Faithful Model Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-12-15 paper/appendix.tex
rerendered appendix

To enable others to easily replicate our experiments, we have chosen to work with small neural network architectures and randomly undersampled the \textit{MNIST} dataset (maintaining class balance). All of our final benchmarks could then be run locally on a personal machine, but grid searches and the sheer number of datasets required us to move to high-performance computing clusters to conduct our experiments efficiently.

\subsubsection{Local runs}

The longest runtimes we experienced for model training and counterfactual benchmarking for a single benchmark were on the order of 12-24 hours (\textit{MNIST} data). For the synthetic data, single benchmarks could be completed in less than an hour. We have summarised our system information below:

\textbf{Software}:

\begin{itemize}
  \item System Version: macOS 13.3.1
  \item Kernel Version: Darwin 22.4.0
\end{itemize}

\textbf{Hardware}:

\begin{itemize}
  \item Model Name: MacBook Pro
  \item Model Identifier: MacBookPro16,1
  \item Processor Name: 8-Core Intel Core i9
  \item Processor Speed: 2.3 GHz
  \item Number of Processors: 1
  \item Total Number of Cores: 8
  \item L2 Cache (per Core): 256 KB
  \item L3 Cache: 16 MB
  \item Hyper-Threading Technology: Enabled
  \item Memory: 32 GB
\end{itemize}

\subsubsection{HPC}

We used two large clusters, which we do not cite or mention by name at this point to not interfer with the double-blind review process. All of our grid searches were multi-processed on 100 CPUs at 4-8GB or memory each.

>>>>

Research reported in this work was partially or completely facilitated by computational resources and support of the DelftBlue~\citep{DHPC2022} and the Delft AI Cluster (DAIC: https://doc.daic.tudelft.nl/) at TU Delft.

For grid search, we used 300 CPUs for tabular real-world datasets ($<$1.5 hours each), 150 CPUs for synthetic datasets ($\approx$1 hour each) and 100 CPUs for vision datasets ($<$4 hours each), where we ran a smaller grid search for the latter. To generate the final results reported in the tables we used 300 CPUs for all datasets ($<$1.5 hours each) except the vision datasets. For the latter, we used 50 CPUs for smaller final experiments and at longer run times ($\approx$5 hours each).

### 2023-12-15 paper/camera-ready/altmeyer.tex
rerendered appendix

The first penalty term involving $\lambda_1$ induces closeness like in~\citet{wachter2017counterfactual}. The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual. The third and final penalty term involving $\lambda_3$ ensures that the generated counterfactual is associated with low predictive uncertainty. These hyperparameters have been tuned through grid search.

@@

Full details concerning model training as well as detailed descriptions and results for all datasets can be found in the appendix. In the following, we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}. The tables show sample averages along with standard deviations across multiple runs for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}). For each metric, the best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). For the tabular datasets, we use the default Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively. The third metric presented~\ref{tab:results-tabular} in Table quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}. For vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.

>>>>

The first penalty term involving $\lambda_1$ induces closeness like in~\citet{wachter2017counterfactual}. The second penalty term involving $\lambda_2$ induces faithfulness by constraining the energy of the generated counterfactual. The third and final penalty term involving $\lambda_3$ ensures that the generated counterfactual is associated with low predictive uncertainty. To tune theses hyperparameters we have relied on grid search.

@@

Full details concerning model training as well as detailed descriptions and results for all datasets can be found in the appendix. In the following, we will focus on the most relevant results highlighted in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision}. The tables show sample averages along with standard deviations across multiple runs for our key evaluation metrics for the \textit{California Housing} and \textit{GMSC} datasets (Table~\ref{tab:results-tabular}) and the \textit{MNIST} dataset (Table~\ref{tab:results-vision}). For each metric, the best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). For the tabular datasets, we use the default Euclidian distance to measure unfaithfulness and implausibility as defined in Equations~\ref{eq:faith} and~\ref{eq:impl}, respectively. The third metric presented~\ref{tab:results-tabular} in Table quantifies the predictive uncertainty of the counterfactual as measured by Equation~\ref{eq:setsize}. For the vision datasets, we rely on measuring the structural dissimilarity between images for our unfaithfulness and implausibility metrics~\citep{wang2003multiscale}.
